Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-12-27 article.tex
edits

Predicting the trajectory of a mutation requires (i) significant fitness difference between genomes carrying different variants at the site and (ii) a selection pressure that changes slowly over time.
Under such conditions, it is expected that frequency trajectories will show a persistent behavior which would make them predictable for some time.
However, we could find only limited evidence for such persistent behavior in the past 19 years of IAV evolution.
\pierre{We could change point (i) below to only mention A/H3N2. }
This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) many methods to predict influenza evolution are unable to predict future frequency dynamics even though they work well for predicting the composition of the future population.
This is illustrated by the case of LBI, which works primarily because it picks strains that represent the future well and not because it is predictive of future dynamics.
% work primarily because they pick strains that represent the future well, not because they predict future dynamics.

The primary focus in this work was the investigation of frequency trajectories of new amino acid mutations.
In the short term, we found that on average the direction of trajectory does not persist for longer than a few months.
Indeed, the average trajectory in figure \ref{fig:trajectories} takes a sharp turn when going from $t<0$ to $t>0$, instead of showing ``inertia''.
This suggests that selective sweeps are not representative of typical trajectories. \pierre{We could cite the Illingworth paper here. They do show that as well, in a different way.}

On a longer timescale, we investigated the probability that a novel mutation observed at frequency $f$ fixes.
In neutral models of evolution this probability equals $f$, while it should be higher or lower than $f$ for mutations with a beneficial or deleterious effect on fitness, respectively.
However, in the case of influenza, this probability differs little from $f$, making current frequency the best predictor for fixation.
In figures~\ref{fig:pfix_stratified} and~\sref{fig:pfix_v_freq_indivjohnsmodels} , we split trajectories into groups for which we expected $P_{fix}$ to deviate from $f$.
Many of these splits, such as high/low LBI or epitope/non-epitope positions, did not result in an increased predictability, while others gave limited information on fixation.
Despite the lack of predictability of mutation frequency trajectories, influenza surface proteins show strong signatures of selection \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.

@@

At the same time, influenza virus phylogenies show clear deviations from those expected from the neutral Kingman coalescent, similar to those expected under Bolthausen-Sznitman coalescent (BSC) processes that are generated by traveling wave models of rapid evolution \citep{neher_genealogies_2013,desai_genetic_2013}.
The correspondence between the BSC and traveling wave models comes from transient exponential amplification of fit strains before these fitness differences are wiped out by further mutation.
This exponential amplification generates long-tailed effective offspring distributions which in turn can leads to genealogies described by the BSC \citep{schweinsberg_coalescent_2003,neher_genealogies_2013}.
Many processes other than selection, including seasonality and spatio-temporal heterogeneity, can generate effective long tailed offspring distributions even in absence of bona-fide fitness differences, which might explain ladder-like non-Kingman phylogenetic trees.

>>>>

Predicting the trajectory of a mutation requires (i) significant fitness difference between genomes carrying different variants at the site and (ii) a selection pressure that changes slowly over time.
Under such conditions, it is expected that frequency trajectories will show a persistent behavior which would make them predictable for some time.
However, we could find little evidence for such persistent behavior in the past 19 years of A/H3N2 evolution.
This lead us to conclude that (i) A/H3N2 influenza virus evolution is qualitatively different from models of rapidly adapting populations (despite clear evidence for frequent positive selection), and (ii) many methods to predict influenza evolution are unable to predict future frequency dynamics even though they work well for predicting the composition of the future population.
This is illustrated by the case of LBI, which works primarily because it picks strains that represent the future well and not because it is predictive of future dynamics.
A/H1N1pdm, in contrast, shows stronger signatures of predictability.

The primary focus in this work was the investigation of frequency trajectories of new amino acid mutations.
In the short term, we found that on average the direction of trajectory does not persist for longer than a few months.
Indeed, the average trajectory in figure \ref{fig:trajectories} takes a sharp turn when going from $t<0$ to $t>0$, instead of showing ``inertia''.
This suggests that selective sweeps are not representative of typical trajectories \citep{illingworth_components_2012}.

On a longer timescale, we investigated the probability that a novel mutation observed at frequency $f$ fixes.
In neutral models of evolution this probability equals $f$, while it should be higher or lower than $f$ for mutations with a beneficial or deleterious effect on fitness, respectively.
However, in the case of influenza, this probability differs little from $f$, making current frequency the best predictor for fixation.
In figures~\ref{fig:pfix_stratified} and~\sref{fig:pfix_v_freq_indivjohnsmodels}, we split trajectories into groups for which we expected $P_{fix}$ to deviate from $f$.
Many of these splits, such as high/low LBI or epitope/non-epitope positions, did not result in an increased predictability, while others gave limited information on fixation.

@@

At the same time, influenza virus surface proteins show strong signatures of selection \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
Their phylogenies show clear deviations from those expected from the neutral Kingman coalescent, similar to those expected under Bolthausen-Sznitman coalescent (BSC) processes that are generated by traveling wave models of rapid evolution \citep{neher_genealogies_2013,desai_genetic_2013}.
The correspondence between the BSC and traveling wave models comes from transient exponential amplification of fit strains before these fitness differences are wiped out by further mutation.
This exponential amplification generates long-tailed effective offspring distributions which in turn can leads to genealogies described by the BSC \citep{schweinsberg_coalescent_2003,neher_genealogies_2013}.
Many processes other than selection, including seasonality and spatio-temporal heterogeneity, can generate effective long tailed offspring distributions even in absence of bona-fide fitness differences, which might explain ladder-like non-Kingman phylogenetic trees.
